---
title: "Lecture 5: Point Level Models - Variograms and EDA"
output:
  revealjs::revealjs_presentation:
    theme: night
    center: true
    transition: none
    incremental: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(knitr)
library(ggplot2)
library(dplyr)
library(gstat)
library(sp)
```

# Class Intro

## Intro Questions 
- What is point referenced data?
- How do we predict the response at unobserved locations?
- What is a variogram?
- For Today:
    - How is a variogram useful?
    - How is EDA used for point referenced data?


## Spherical semivariogram: Exercise
$$\gamma(d)=
\begin{cases}
\tau^2 + \sigma^2 \; \; \text{if } d \geq 1/ \phi\\
\tau^2 + \sigma^2\left[\frac{3\phi d}{2} - \frac{1}{2} (\phi d)^3 \right]\\
0 \; \; \text{otherwise}
\end{cases}$$

- Sketch, or generate in R, a spherical semivariogram
- On this figure label the nugget, sill, partial sill, and range.

## Spherical semivariogram: Solution

```{r}
tau.sq <- 1
sigma.sq <- 2
range <- 2
phi <- 1/range
d <- seq(0,3, by =.01)
sph.gam <- tau.sq + sigma.sq * (3 * phi * d / 2 - .5 * (phi*d)^3)
sph.gam[d >= range] <- tau.sq + sigma.sq
sph.var <- data.frame(d=d, sph.gam = sph.gam )
ggplot(data = sph.var, aes(x=d, y=sph.gam)) + geom_line() + ylim(0,4) + ylab('spherical variogram') + annotate('text', x=0.1, y = 0.9, label = 'nugget') + annotate('text', x=2.5, y = 3.1, label = 'sill') + annotate('text', x=2.2, y = .1, label = 'range', color = 'blue') +  annotate("segment", x = 2, xend = 2, y = 0, yend = 4,
  colour = "blue", linetype =3, size = .5) +  annotate("segment", x = 2, xend = 3, y = 1, yend = 1,colour = "red", linetype =3, size = .5) +  annotate("segment", x = 2, xend = 3, y = 3, yend = 3,colour = "red", linetype =3, size = .5) +  annotate("segment", x = 2.5, xend = 2.5, y = 1, yend = 3,colour = "red", linetype =3, size = .5) + annotate('text', x=2.2, y = 2, label = 'partial sill', color = 'red') 
```

# More Variograms

## Exponential 

>- We saw the exponential covariance earlier in class, what is the mathematical form?
- $$C(d)=
\begin{cases}
\tau^2 + \sigma^2 \; \; \text{if } d = 0\\
\sigma^2 \exp(-\phi d) \; \; \text{if } d > 0
\end{cases}$$
- The variogram is 
$$\gamma(d)=
\begin{cases}
\tau^2 + \sigma^2(1 - \exp(-\phi d)) \; \; \text{if } d > 0\\
0 \; \; \text{otherwise} 
\end{cases}$$

## Exponential Semivariogram: Exercise

- Plot the exponential semivariogram

- Identify and interpret: the nugget, range, and sill. How are they similar or different from the spherical semivariogram?

## Exponential Semivariogram: Solution

```{r}
exp.gam <- tau.sq + sigma.sq * (1 - exp( - phi * d))
exp.var <- data.frame(d=d, exp.gam = exp.gam )
ggplot(data = exp.var, aes(x=d, y=exp.gam)) + geom_line() + ylim(0,4) + ylab('exponential variogram') + annotate('text', x=0.1, y = 0.9, label = 'nugget') + annotate('text', x=2.9, y = 2.65, label = 'sill?') + annotate('text', x=2.2, y = .1, label = 'range?', color = 'blue') +  annotate("segment", x = 2, xend = 2, y = 0, yend = 4,
  colour = "blue", linetype =3, size = .5) 
```

## Exponential Semivariogram: Solution 2

- The sill is only reached asymptotically when $$\lim_{d\rightarrow \infty} exp(-\phi * d) \rightarrow 0$$
- Thus, the range, defined as distance where semivariogram reaches the sill is $\infty$
- The *effective range* is defined as the distance where there is *effectively* no spatial structure. Generally this is determined when the correlation is .05. 
- \begin{eqnarray*}
\exp(-d_o \phi) &=&.05 \\
d_0 \phi &=& - \log(.05) \\
d_0 &\approx& \frac{3}{\phi}
\end{eqnarray*}

## Exponential Semivariogram: Solution 3

```{r}
d <- seq(0,8, by = .01)
exp.gam <- tau.sq + sigma.sq * (1 - exp( - phi * d))
exp.var <- data.frame(d=d, exp.gam = exp.gam )
ggplot(data = exp.var, aes(x=d, y=exp.gam)) + geom_line() + ylim(0,4) + ylab('exponential variogram') + annotate('text', x=0.1, y = 0.9, label = 'nugget') + annotate('text', x=7.9, y = 3.1, label = 'sill') + annotate('text', x=6.6, y = .1, label = 'effective range', color = 'blue') +  annotate("segment", x = 6, xend = 6, y = 0, yend = 4,
  colour = "blue", linetype =3, size = .5) 
```

## More Semivariograms: Equations
- Gaussian: 
$$\gamma(d)=
\begin{cases}
\tau^2 + \sigma^2(1 - \exp(-\phi^2 d^2)) \; \; \text{if } d > 0\\
0 \; \; \text{otherwise} 
\end{cases}$$
- Powered Exponential:
$$\gamma(d)=
\begin{cases}
\tau^2 + \sigma^2(1 - \exp(-|\phi d|^p)) \; \; \text{if } d > 0\\
0 \; \; \text{otherwise} 
\end{cases}$$
- Mat$\acute{\text{e}}$rn:
$$\gamma(d)=
\begin{cases}
\tau^2 + \sigma^2\left[1 - \frac{(2\sqrt{\nu}d \phi)^\nu}{2^{\nu-1}\Gamma(\nu)} K_\nu (2 \sqrt{\nu} d \phi)\right] \; \; \text{if } d > 0\\
0 \; \; \text{otherwise} 
\end{cases},$$
where $K_\nu$ is a modified Bessel function and $\Gamma()$ is a Gamma function.


# Variogram Creation


## Variogram Creation: How?

```{r, echo=F}
data(meuse)
meuse.small <- meuse %>% select(x, y, copper) %>% tbl_df()
meuse.small %>% head(15) %>% kable()
```

## Variogram Creation: Steps

1. Calculate distances between sampling locations
2. Choose grid for distance calculations
3. Calculate empirical semivariogram
$$\hat{\gamma}(d_k) = \frac{1}{2N(d_k)} \sum_{\boldsymbol{s}_i,\boldsymbol{s}_i, \in N(d_k)}\left[ \boldsymbol{Y}(\boldsymbol{s}_i) - \boldsymbol{Y}(\boldsymbol{s}_j)  \right]^2,$$
where $$N(d_k) = \{(\boldsymbol{s}_i,\boldsymbol{s}_j): ||\boldsymbol{s}_i -\boldsymbol{s}_j|| \in I_k\}$$ and $I_k$ is the $k^{th}$ interval.
4. Plot the semivariogram


## Variogram Creation: R function

```{r, eval = T}
data(meuse)
coordinates(meuse) = ~x+y
variogram(copper~1, meuse) %>% plot()
```

## Variogram Creation: How - Step 1

Calculate Distances between sampling locations

```{r, echo= t}
dist(meuse.small)
dist.mat <- dist(meuse.small %>% select(x,y))
```

## Variogram Creation: How - Step 2

Choose grid for distance calculations

```{r, echo= t}
cutoff <- max(dist.mat) / 3 # default maximum distance
num.bins <-  15
bin.width <- cutoff / 15
```

## Variogram Creation: How - Step 3

Calculate empirical semivariogram

